---
title: "Evaluate gene sets by cell type-specific gene expression"
author: "Wei Sun"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: journal
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
    number_sections: false
  df_print: paged
params:
  cell_type_name: "Astro"
  graph_weight: "2.0" 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width = 80)
```


# Setup

## Load R libraries
```{r load_libraries, message=FALSE, warning=FALSE}
library(data.table)
library(ggplot2)
library(ggpubr)
library(tidyr)
library(limma)
library(stringr)

theme_set(theme_classic())

cell_type_name = params$cell_type_name
graph_weight = params$graph_weight

cell_type_name
graph_weight
```

# Check the expression of each gene set in each brain-related cell type

## Read in gene info and gene set assignments

```{r}
file_tag = sprintf("%s_%s", cell_type_name, graph_weight)

assayed_genes = scan(sprintf("output/gene_list_%s.txt", file_tag), 
                     what = character(), sep="\n")

length(assayed_genes)

gene_sets = scan(sprintf("output/name_s_%s.txt", file_tag), 
                 what = character(), sep="\n")

gene_sets = sapply(gene_sets, strsplit, split=",")
n_genes   = sapply(gene_sets, length)
names(n_genes) = NULL
summary(n_genes)
length(n_genes)
sort(n_genes)
```

### Find gene symbols using the `alias2Symbol` function from `limma`.

```{r}
a2s = rep(NA, length(assayed_genes))
for(i in 1:length(assayed_genes)){
  gi = assayed_genes[i]
  ai = alias2Symbol(gi)
  if(length(ai) > 1){
    print(gi)
    print(ai)
  }
  a2s[i] = ai[1]
}

table(is.na(a2s))

table(a2s == assayed_genes, useNA = 'ifany')

gene_info = data.table(sym_in_data = assayed_genes, sym_limma = a2s)

gene_info[sym_in_data != sym_limma,]
gene_info[, gene_symbol := sym_in_data]
gene_info[which(sym_in_data != sym_limma & (gene_symbol != "MT-CO2")), 
                gene_symbol := sym_limma]

dim(gene_info)
gene_info[1:5,]
t1 = table(gene_info$gene_symbol)
table(t1)
```

## Read in cell type-specific expression data

Gene set annotations (by gene symbols) were downloaded from MSigDB website. 

```{r}
ct = fread("../Annotation/rna_single_cell_type.tsv.gz")
dim(ct)
ct[1:5,]
length(unique(ct$`Cell type`))
table(ct$`Cell type`)

cell_group = fread("../Annotation/rna_single_cell_type_brain.tsv")
dim(cell_group)
cell_group[1:2,]

ct = ct[`Cell type` %in% cell_group$Cell_type,]
table(ct$`Cell type`)
```

## Check gene expression for each gene set

```{r warning = FALSE, message = FALSE}
dim(gene_info)

for(k in 1:length(gene_sets)){
  if(length(gene_sets[[k]]) < 10) { next }
  
  print(k)
  set_k = paste0("set_", k)
  print(gene_sets[[k]])
  
  genes = gene_info[sym_in_data %in% gene_sets[[k]], gene_symbol]

  n_genes = sum(genes %in% ct$`Gene name`)
  print(sprintf("found %d genes.", n_genes))

  if(n_genes == 0) { next }

  df = ct[`Gene name` %in% genes,]
  dim(df)
  df[1:2,]
  
  stopifnot(all(str_to_lower(df$`Cell type`) %in% 
                  str_to_lower(cell_group$Cell_type)))
  
  mat1 = match(str_to_lower(df$`Cell type`), 
               str_to_lower(cell_group$Cell_type))
  
  df = cbind(df, cell_group[mat1,])
  df[1:2,]
  
  df$Cell_type = factor(df$Cell_type, levels = cell_group$Cell_type)
  df$Cell_group = factor(df$Cell_group, 
                         levels = unique(cell_group$Cell_group))
  
  p1 = ggplot(df, aes(x=Cell_type, y=log10(nTPM + 0.1), fill=Cell_group)) + 
    geom_boxplot() + xlab("Cell type") + ggtitle(set_k) + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
    scale_fill_brewer(palette="RdBu")

  print(p1)
}

```



# Session information
```{r}
gc()

sessionInfo()
```

