---
title: "Check genes associated with case-control status"
author: "Wei Sun"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: journal
    highlight: tango
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: false
    number_sections: false
  df_print: paged
params:
  graph_weight: "2.0" 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(width = 80)
```


# Setup

## Load R libraries
```{r load_libraries, message=FALSE, warning=FALSE}
library(data.table)
library(ggplot2)
library(ggpubr)
library(tidyr)
library(limma)
library(stringr)
library(Matrix)
library(gplots)
library(zellkonverter)
library(pROC)
library(doParallel)
library(foreach)
library(SummarizedExperiment)
library(reticulate)

theme_set(theme_classic())

cell_type_name = "CD8T"
graph_weight = params$graph_weight
graph_weight

nCore = 10

registerDoParallel(cores=nCore)
options(mc.cores=nCore)
getOption("mc.cores")

np = import("numpy")
```

# Examine results

## Check assignment matrix S

```{r fig.width=3, fig.height=3}
file_tag = sprintf("%s_%s", cell_type_name, graph_weight)

s_mat = np$load(sprintf("output/train_s_%s.npy", file_tag))
dim(s_mat)
s_mat = data.matrix(s_mat)
dim(s_mat)
s_mat[1:2,1:3]

colSums(s_mat > 0.8)

v_s = c(s_mat)
summary(v_s)
table(v_s <= 0.1)
mean(v_s <= 0.1)

df1 = data.frame(score = v_s[v_s > 0.1])
ggplot(data = df1, aes(x = score)) +
  geom_histogram(breaks = seq(0.1, 1, by=0.05), fill = "skyblue", 
                 color = "darkblue", alpha = 0.7)
```


# Check association using gene expression data

## Read in gene and cell info
```{r fig.width=3, fig.height=3}
genes2use = scan(sprintf("output/gene_list_%s.txt", file_tag), 
             what=character())
length(genes2use)
genes2use[1:5]

cell_info = fread("data/yost_cd8_meta.tsv")
dim(cell_info)
cell_info[1:2,]
```

## Read in reduced representation

```{r fig.width=3, fig.height=3}
test_reduced = readH5AD(sprintf("output/test_reduced_%s.h5ad", file_tag))
test_reduced

colData(test_reduced)

rdat = assay(test_reduced, "X")
dim(rdat)

rdat[1:2,1:3]

cell_info2 = data.frame(colData(test_reduced))
dim(cell_info2)
cell_info2[1:2,]

cell_info = cbind(cell_info, cell_info2)
dim(cell_info)
cell_info[1:2,]

names(cell_info)[names(cell_info) == "cell_type"] = "response"
```

## Read in the complete gene expression data 

Subset data by genes and cells. Since Yost 2019 data are used in testing data, some genes in the gene set (from training data of SF 2018) are not included in the gene expression. 

```{r fig.width=4.5, fig.height=3}
edat = fread("data/yost_cd8_counts.tsv.gz")
dim(edat)
edat[1:2,1:3]

genes_all = edat$gene

edat = data.matrix(edat[,-1])
dim(edat)
edat[1:2,1:3]

stopifnot(all(cell_info$cell.id == colnames(edat)))
cell_info$rd = colSums(edat)

table(genes2use %in% genes_all)
w2use = which(genes2use %in% genes_all)
s_mat = s_mat[w2use,]
dim(s_mat)
genes2use = genes2use[w2use]

edat = edat[match(genes2use, genes_all),]
dim(edat)

ggplot(data = cell_info, aes(x = rd)) +
  geom_histogram(bins=20, fill = "skyblue", color = "darkblue", alpha = 0.7)

ggplot(data = cell_info, aes(x = rd)) + 
  geom_density(aes(color=response, fill=response), alpha = 0.5)

ggplot(data = cell_info, aes(x = log10(rd))) + 
  geom_density(aes(color=response, fill=response), alpha = 0.5)
```

## Evaluate AUC for each of the selected genes.

```{r fig.width=8, fig.height=3}
y_true = cell_info$response
table(y_true)

auc_vals = foreach(i = 1:nrow(edat), .combine = "c") %dopar% {
  y1 = edat[i,]
  auc1 = auc(roc(y_true, y1, levels=c("No", "Yes"), direction = "<"))
  as.numeric(auc1)
}

length(auc_vals)
summary(auc_vals)

df1 = data.frame(gene=genes2use, auc=auc_vals, gene_set = 0)

for(k in 1:ncol(s_mat)){
  genes_k = which(s_mat[,k] > 0.8)
  if(length(genes_k) > 0) df1$gene_set[genes_k] = k
}

tapply(df1$auc, df1$gene_set==0, summary)

df1$gene_set[df1$gene_set == 0] = "B"
df1$gene_set = factor(df1$gene_set, levels = as.character(0:40))

ggplot(df1, aes(x = gene_set, y = auc, color = gene_set)) +
  geom_boxplot(show.legend = FALSE) +
  labs(x = "Gene sets", y = "AUC") + 
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red")
```


## Check the AUC of each reduced component and its distribution. 

```{r fig.width=5, fig.height=3}
y_true = cell_info$response
table(y_true)

auc_r = foreach(i = 1:nrow(rdat), .combine = "c") %dopar% {
  y1 = rdat[i,]
  auc1 = auc(roc(y_true, y1, levels=c("No", "Yes"), 
                 direction = "<"))
  as.numeric(auc1)
}
summary(auc_r)

stopifnot(nrow(rdat) == 40)

df_r = data.frame(component = 1:40, auc=auc_r, n_genes = colSums(s_mat > 0.8))

ggplot(data = df_r, aes(x=component, y=auc)) + 
  geom_point() + geom_text(data = subset(df_r, n_genes > 0),
                  aes(label = n_genes), hjust = -0.5) + 
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "red") + 
  geom_vline(xintercept = seq(5,40,by=5), color = "grey")

ggplot(df_r, aes(x = auc)) +
  geom_histogram(aes(y = after_stat(density)), binwidth = 0.05, 
                 fill = "lightblue", color = "black", alpha = 0.7) + 
  geom_density(data=df1, aes(x = auc), color = "red", linewidth = 1)

```


# Session information

```{r}
gc()

sessionInfo()
```

